Supplementary figures\nCapturing regeneration in models of forest dynamics

### 2023-03-12 manuscript/manuscriptRW.Rmd
combine the 3 gradients figure in one figure and add the zoomed ones into the appendix

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @langan2017, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2004, @morin2021, @lasch-born2020, @hickler2012, @fischer_lessons_2016
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Two models, ForClim and ForCEEPS, were used here with two alternative
variants. ForClim variant 1 [@bugmann1996] is based on a recruitment
module that adheres closely to the concept introduced by @botkin1972,
whereas ForClim variant 11 is adopting a rather different approach that
considers space filling along with different 'points of entry' for the
action of specific environmental factors [@huber2020]. These two model
variants allow us to evaluate a more process-based module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback (i.e. simulated actual
composition and relative abundance of species in the plot)via seed trees
for the quantity and quality of simulated ingrowth, as one variant
includes this feedback whereas the other does not.

@@

The simulation results were analyzed regarding the ingrowth levels
(i.e., ingrowth number per unit time and space), species composition of
the ingrowth, and ingrowth mortality. The observed data were modeled
using a Generalized Additive Model [@wood2011] with a negative binomial
distribution and restricted maximum likelihood to better understand the
environmental gradients trend in the levels of the observed ingrowth,
relative to the simulation results per model.

Ingrowth levels were evaluated along three key environmental gradients,
i.e., stand basal area (the basal area of all the trees in each sample
including tree establishment) as a proxy of light availability at the
forest floor, the annual degree-day sum (DD) [@allen1976; @fischlin1995]
as a proxy for growing season warmth, and the climatic water balance
(CWB) as a proxy for soil moisture [@speich2019]. The mean ingrowth
values were calculated as the mean across the 200 samples per site.

We evaluated the species diversity across models and in relation to the
observed data by calculating the Shannon index based on the relative
proportion of the species in terms of basal area. The Shannon index was
calculated for the establishment ($H_{n,r}$) (Equation
\ref{eqpiShannonR}) and at the stand level ($H_{n,s}$)(Equation
\ref{eqpiShannon}). The higher the value of the index, the higher the
species diversity at a particular site and sample.

We also tested for significant trends across XXXXX !!!!!

@@

In order to analyze the relationship between the adult species
composition and ingrowth across the climatic gradients, we focused on
the five main tree species or genera: *Abies alba*, *Fagus sylvatica*,
*Picea abies*, *Pinus sylvestris*, and *Quercus* spp. For these species,
we calculated the mean ingrowth basal area share per site as the mean
value across the available samples per site:

@@

```{=tex}
\begin{equation}
Reineke = (7/10)^{(-1.605)} \label{eqReineke}
\end{equation}
```
# Results

The ingrowth levels varied strongly across the 15 models and typically
did not match the levels found in the forest reserve data (Figure
\@ref(fig:over)). Ingrowth was overestimated in most models for both the
7 and 10 cm diameter thresholds, with the exception of the empirical
stand model SIBYLA, the landscape model Landis-II and the global model
aDGVM2, which estimated ingrowth rates at the lower end of the
plausibility interval of the observed data (Figure \@ref(fig:over)). The
models with the largest overestimation were the stand model PICUS and
the landscape model TreeMig. For most models, the variability of
simulated ingrowth levels across the 200 sites (visible from the
interquartile range in the box plots of Figure \@ref(fig:over)) was
similar to or smaller than observed, with the notable exception of
PICUS, where simulated ingrowth variability was much larger.

```{r over, echo=FALSE, fig.cap="Mean ingrowth levels across all samples per site, plotted for the 200 sites and for each model. The red dashed lines shows the 25th and 75th percentiles for the 7 cm diameter threshold in the observed data. There are two boxplots for each model where the left lighted boxplot corresponds to 7cm and the right darker boxplot to 10cm", out.width='100%'}
        knitr::include_graphics(here::here("figures", "recOverAll710.png"))
```

@@

Both observations and simulated data had no ingrowth in some samples and
at some sites (Table S1). The observed data had 3% of the samples with
no ingrowth. Only three models had a larger proportion of no ingrowth
(4C, Landis-II and aDVM2). CHECK! No LPJ-GUESS anymore with the new data
submission!!!! Two models always simulated ingrowth for both the 7 and
10 cm threshold (TreeMig and LPJ-GUESS), i.e., they did not feature any
zero values. The other models had a very low percentage of samples with
no ingrowth (0.01% - 1.31%), i.e., they had distinctly fewer occurrences
of zeros compared to the observations. There was no significant
relationship between the mean complexity of the establishment module
(Table \@ref(tab:tableModels)) and the overestimation proportion at a
diameter threshold of 7 cm (SM3).

```{r nvTrend710BA, echo=FALSE, fig.cap="Mean ingrowth values across the 200 samples per site, for the 200 sites against the stand basal area. (A) same y-axis across all models; (B) y-axis scaled by model. The values were split into ten bins; the red lines represent a GAM model showing the trend in the observed data.", fig.show = 'hold', out.width = '80%'}
    knitr::include_graphics(c(here::here("figures", "envTrend7Totba_500.png"),
                              here::here("figures", "envTrend7Totba.png")))
```

Most models matched the level of diversity of the observed data quite
well as quantified by the Shannon index (Figure \@ref(fig:H710)),
although they consistently underestimated its variability across sites.
Four models overestimated ingrowth diversity: 4C, PICUS, Landis-II and
LPJ-GUESS, the latter mostly for the 10 cm diameter threshold. The model
4C is a special case, as it simulated five species only, i.e., its
diversity values are not directly comparable to those of the other
models, nor to the observations. Only one model, SIBYLA, consistently
underestimated ingrowth diversity.

In most models, there were only small but significant differences in the
species diversity of ingrowth between 7 and 10 cm across sites (Tables
S2-S4). Three models (4C, SIBYLA, Landis-II) maintained the ingrowth
diversity between the 7 and 10 cm thresholds (i.e., the differences
between them were not significant, see Table S3), and the same was
evident from the observed data. There were a few models with a strong
increase of ingrowth diversity between 7 and 10 cm (ForCEEPS, TreeMig,
LPJ-GUESS, FORMIND, Figure \@ref(fig:H710)).

For both the observed and the simulated data, and both ingrowth
thresholds, species diversity was lower in the ingrowth compared to
stand-level diversity (Figures S3 and S4). Yet, models simulating higher
diversity at the stand level featured higher diversity in ingrowth
(Figure S3 and S4). Most models captured reasonably well or
overpredicted species diversity at the stand level (Figure
\@ref(fig:combinedRichness)A, center; cf. Figure S5), but a
characteristic feature was that simulations had a much lower variability
of diversity compared to observations. Models overpredicting species
diversity included ForClim1, ForClim 11, FORMIND and TreeMig (Figure
\@ref(fig:combinedRichness)A, left). Several models underpredicted
species diversity, i.e., LPJ-GUESS, ForCEEPS(f) and 4C (Figure
\@ref(fig:combinedRichness)A, right). The majority of the models
captured reasonably well the ingrowth diversity at lower values of
diversity, but most of them strongly underestimated ingrowth diversity
for higher observed diversity levels (Figure
\@ref(fig:combinedRichness)B, cf. Figure S6). Both the models with
feedback and those without feedback from the adult trees to ingrowth
(via seed production) underestimated ingrowth species diversity compared
to the observations. Again, simulated diversity was considerably less
variable than observed diversity (Fig. \@ref(fig:combinedRichness)B).

@@

There were strong differences among the models regarding the mortality
rate between 7 and 10 cm (Figure \@ref(fig:mortality11)). Most of the
models featured a mortality rate significantly larger than the observed
data, thus at least partially compensating for the general
overestimation of ingrowth levels (Figure \@ref(fig:over) and Table S5);
six models did not feature a significant difference compared to the
observed data (ForCEEPS, FORMIND, ForClim 1 and 11, and aDGVM2), and
only the model LPJ-GUESS had a significantly lower mortality rate. Four
models (xComp, PICUS, LandClim and TreeMig) featured very high mortality
rates (i.e., well above the Reineke self-thinning line), which
compensated for the strong initial overestimation of ingrowth (cf.
Figure 2). Interestingly, two models that underestimated overall
ingrowth levels (SIBYLA and Landis-II; cf. Figure \@ref(fig:over))
featured mortality rates that were close to but still above the
self-thinning line (Figure \@ref(fig:mortality11)).

Lastly, some models that had a low number of individuals at 7 cm
threshold also had a smaller decrease of ingrowth between 7 and 10 cm
(ratio 7-10 cm) (Figure S10). In these models, competition and
self-thinning are either not pronounced or must have occurred before the
species had reached 7 cm. However, this pattern was not consistent
across models (Figure S10). It is noteworthy that there is a
relationship between the ingrowth rate at 7 cm and the mortality rate:
most of the models showed a positive effect (higher ingrowth at 7 cm is
coupled to an increase of the mortality between 7 and 10 cm). While some
models showed a negative effect, this linear trend was not pronounced
and not always significant regarding the slope (Figure S9 and Table S6).

>>>>

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
    extra_dependencies: "subfig"
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @langan2017, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2004, @morin2021, @lasch-born2020, @hickler2012, @fischer_lessons_2016
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Two models, ForClim and ForCEEPS, were used here with two alternative
variants. ForClim variant 1 [@bugmann1996] is based on a recruitment
module that adheres closely to the concept introduced by @botkin1972,
whereas ForClim variant 11 is adopting a rather different approach that
considers space filling along with different 'points of entry' for the
action of specific environmental factors [@huber2020]. These two model
variants allow us to evaluate a more process-based module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback (i.e. simulated actual
composition and relative abundance of species in the plot) via seed
trees for the quantity and quality of simulated ingrowth, as one variant
includes this feedback whereas the other does not.

@@

The simulation results were analyzed regarding the ingrowth levels
(i.e., ingrowth number per unit time and space), species diversity of
the ingrowth, and ingrowth mortality. We evaluated the species diversity
across models and in relation to the observed data by calculating the
Shannon index based on the relative proportion of the species in terms
of basal area. The Shannon index was calculated for the establishment
($H_{n,r}$) (Equation \ref{eqpiShannonR}) and at the stand level
($H_{n,s}$)(Equation \ref{eqpiShannon}). The higher the value of the
index, the higher the species diversity at a particular site and sample.

@@

```{=tex}
\begin{equation}
Reineke = (7/10)^{(-1.605)} \label{eqReineke}
\end{equation}
```
We assessed model traits in relation to model performance focusing on
features such as model complexity as defined in @bugmann2022, model type
(empirical or process based), models with and without canopy feedback
(the level and species composition of the ingrowth is influenced by the
existing adult tree community) and the models scale (stand, landscape or
global). We tested for significant differences considering each of these
model features and between observed and simulated ingrowth
overestimation, ingrowth levels and species diversity respectively.

The total ingrowth and individual species regeneration niches were
evaluated across environmental gradients of light availability,
temperature, and soil moisture. Ingrowth levels were evaluated along
three key environmental gradients, i.e., stand basal area (the basal
area of all the trees in each sample including tree establishment) as a
proxy of light availability at the forest floor, the annual degree-day
sum (DD) [@allen1976; @fischlin1995] as a proxy for growing season
warmth, and the climatic water balance (CWB) as a proxy for soil
moisture [@speich2019]. The mean ingrowth values were calculated as the
mean across the 200 samples per site. The observed data were modeled
using a Generalized Additive Model [@wood2011] with a negative binomial
distribution and restricted maximum likelihood to better understand the
environmental gradients trend in the levels of the observed ingrowth,
relative to the simulation results per model.

In order to analyze the individual species regeneration niches across
the climatic gradients, we focused on the five main tree species or
genera: *Abies alba*, *Fagus sylvatica*, *Picea abies*, *Pinus
sylvestris*, and *Quercus* spp. For these species, we calculated the
mean ingrowth basal area share per site as the mean value across the
available samples per site:

@@

# Results

## Ingrowth levels, tree diversity and mortality in tree establishment

The ingrowth levels varied strongly across the 15 models and typically
did not match the levels found in the forest reserve data (Figure
\@ref(fig:over)). Ingrowth was overestimated in most models for both the
7 and 10 cm diameter thresholds, with the exception of the empirical
stand model SIBYLA, the landscape model Landis-II and the global model
aDGVM2, which estimated ingrowth levels at the lower end of the
plausibility interval of the observed data (Figure \@ref(fig:over)). The
models with the largest overestimation were the stand model PICUS and
the landscape model TreeMig. For most models, the variability of
simulated ingrowth levels across the 200 sites (visible from the
interquartile range in the box plots of Figure \@ref(fig:over)) was
similar to or smaller than observed, with the notable exception of
PICUS, where simulated ingrowth variability was much larger.

```{r over, echo=FALSE, fig.cap="Mean ingrowth levels across all samples per site, plotted for the 200 sites and for each model. The red dashed lines shows the 25th and 75th percentiles for the 7cm diameter threshold in the observed data. There are two boxplots for each model where the left lighted boxplot corresponds to 7cm and the right darker boxplot to 10cm", out.width='100%'}
        knitr::include_graphics(here::here("figures", "recOverAll710.png"))
```

Both observations and simulated data had no ingrowth in some samples and
at some sites (Table S1). The observed data had 4% of the samples with
no ingrowth. Only three models had a larger proportion of no ingrowth
(4C, Landis-II and aDVM2). Two models always simulated ingrowth for both
the 7 and 10 cm threshold (xComp and TreeMig), i.e., they did not
feature any zero values. The other models had a very low percentage of
samples with no ingrowth (0.01% - 2.39%), i.e., they had distinctly
fewer occurrences of zeros compared to the observations.

Most models matched the level of diversity of the observed data quite
well as quantified by the Shannon index (Figure \@ref(fig:H710)). Five
models overestimated ingrowth diversity: ForCEEPS, ForClim 11, PICUS,
TreeMig and LandClim, the latter mostly for the 7 cm diameter threshold.
The model 4C is a special case, as it simulated five species only, i.e.,
its diversity values are not directly comparable to those of the other
models, nor to the observations. Only one model, Landis-II, consistently
underestimated ingrowth diversity.

In most models, there were only small but significant differences in the
species diversity of ingrowth between 7 and 10 cm across sites (Tables
S2). Four models (iLand, Landis-II, TreeMig and LPJ-GUESS) maintained
the ingrowth diversity between the 7 and 10 cm thresholds (i.e., the
differences between them were not significant, see Table S2), and the
same was evident from the observed data.

For both the observed and the simulated data, and both ingrowth
thresholds, species diversity in the ingrowth compared to stand-level
diversity did not have clear patterns across all the models (Figures S1
and S2). Most models captured reasonably well or overpredicted species
diversity at the stand level (Figure \@ref(fig:combinedRichness)A,
center; cf. Figure S3), but a characteristic feature was that
simulations had a much lower variability of diversity compared to
observations. Models overpredicting species diversity included ForClim1,
ForClim 11, FORMIND and TreeMig (Figure \@ref(fig:combinedRichness)A,
left). Several models underpredicted species diversity, i.e., LPJ-GUESS,
ForCEEPS(f) and 4C (Figure \@ref(fig:combinedRichness)A, right).

The majority of the models overpredicted ingrowth diversity and only two
of them underestimated ingrowth diversity (4C and Landis-II) (Figure
\@ref(fig:combinedRichness)A and C, cf. Figure S4). Both the models with
feedback and those without feedback from the adult trees to ingrowth
(via seed production) overestimate, underestimated and captured
reasonably well ingrowth species diversity compared to the observations.

There were strong differences among the models regarding the mortality
rate between 7 and 10 cm (Figure \@ref(fig:mortality11)). Most of the
models featured a mortality rate significantly larger than the observed
data, thus at least partially compensating for the general
overestimation of ingrowth levels (Figure \@ref(fig:over) and Table S5);
six models did not feature a significant difference compared to the
observed data (ForCEEPS, FORMIND, ForClim 1 and 11, iLand and aDGVM2).
Four models (xComp, PICUS, LandClim and TreeMig) featured very high
mortality rates (i.e., well above the Reineke self-thinning line), which
compensated for the strong initial overestimation of ingrowth (cf.
Figure \@ref(fig:over)). Interestingly, two models that underestimated
overall ingrowth levels (SIBYLA and Landis-II; cf. Figure
\@ref(fig:over)) featured mortality rates that were close to but still
above the self-thinning line (Figure \@ref(fig:mortality11)).

Lastly, some models that had a low number of individuals at 7 cm
threshold also had a smaller decrease of ingrowth between 7 and 10 cm
(ratio 7-10 cm) (Figure S5). In these models, competition and
self-thinning are either not pronounced or must have occurred before the
species had reached 7 cm. However, this pattern was not consistent
across models (Figure S5). It is noteworthy that there is a relationship
between the ingrowth rate at 7 cm and the mortality rate (Table S4):
most of the models showed a positive effect (higher ingrowth at 7 cm is
coupled to an increase of the mortality between 7 and 10 cm). While some
models showed a negative effect, this linear trend was not pronounced
and not always significant regarding the slope.

## Model traits and model performance

There was no significant relationship between the mean complexity of the
establishment module (Table \@ref(tab:tableModels)) and the
overestimation proportion at a diameter threshold of 7 cm (Figure S7).
All the models except 4C had significant differences between the
observed mean ingrowth values and the simulated, therefore it was not
possible to differentiate in terms of performance between models forest
type (empirical or process based) and across models scales (stand,
landscape or global) (Table S5). The ingrowth species diversity values
simulated per models were also assessed againts the observed data to see
if there was a difference between models that include feedback compare
to those that did not include feedback, all the models except
ForCEEPs(f) and FORMIND had significant differences (Table S6). It is
noteworthy the change from overpredicted species diversity in the
ingrowth for the model ForCEEPS to a closer diversity levels when the
recruitment module included feedback (ForCEEPs(f)). Again, simulated
diversity was considerably less variable than observed diversity.

## Total ingrowth and individual species regeneration niches

@@

```{r nvTrend710BAlightTemp, echo=FALSE, fig.cap="Mean ingrowth values across the 200 samples per site, for the 200 sites against gradients of (a) total basal area; (b) climatic water balance; (c) seasonal degree sum. The values were split into ten bins; the red lines represent a GAM model showing the trend in the observed data. The same graphs with y-axis scaled by model as availalbe in Figure S", fig.show = 'hold', out.width="49%",fig.ncol = 2, fig.subcap=c('', '', ''), fig.show='hold', fig.align='left'}

   knitr::include_graphics(c(here::here("figures", "envTrend7Totba_500.png"),
                              here::here("figures", "envTrend7WB_500.png"),
                              here::here("figures", "envTrend7DDS_500.png")))
```

@@

### 2023-03-12 manuscript/supplementaryRW.Rmd
combine the 3 gradients figure in one figure and add the zoomed ones into the appendix

### Mortality

@@

### Model complexity VS overestimation

@@

## Total ingrowth and individual species regeneration niches

```{r envTrend7wb, echo=FALSE, fig.cap="Mean ingrowth values across the 200 samples per site, for the 200 sites against the sites water balance. (A) same y-axis across all models; (B) y-axis scaled by model. The values were split into ten bins; the red lines represent a GAM model showing the trend in the observed data.", fig.show = 'hold', out.width = '80%', fig.pos='H'}
    knitr::include_graphics(c(here::here("figures", "envTrend7WB_500.png"),
                              here::here("figures", "envTrend7WB.png")))
```

```{r envTrend7dds, echo=FALSE, fig.cap="Mean ingrowth values across the 200 samples per site, for the 200 sites against the sites seasonal degree-day sum. (A) same y-axis across all models; (B) y-axis scaled by model. The values were split into ten bins; the red lines represent a GAM model showing the trend in the observed data.", fig.show = 'hold', out.width = '80%', fig.pos='H'}
    knitr::include_graphics(c(here::here("figures", "envTrend7DDS_500.png"),
                              here::here("figures", "envTrend7DDS.png")))
```

>>>>

@@

@@

```{r rtreesDiff, echo=FALSE,  message=FALSE, fig.pos='H'}
library(knitr)

rtreesDiff <- readr::read_csv(here::here("figures", "rtreesDiff.csv"))

kable(rtreesDiff,  format = "latex", caption = "Test of the significance of the differences of rate of recruitment at 7 cm threshold of the observed values and the simulated values per model.") |>
    kableExtra::kable_styling(latex_options = "HOLD_position") |>
    kableExtra::kable_styling(full_width = F, latex_options = "scale_down")
```

```{r sigH7_SimObs, echo=FALSE,  message=FALSE, fig.pos='H'}
library(knitr)

rtreesDiff <- readr::read_csv(here::here("figures", "sigH7_SimObs.csv"))

kable(rtreesDiff,  format = "latex", caption = "Test of the significance of the differences of rate of recruitment at 7 cm threshold of the observed values and the simulated values per model.") |>
    kableExtra::kable_styling(latex_options = "HOLD_position") |>
    kableExtra::kable_styling(full_width = F, latex_options = "scale_down")
```

\newpage
## Total ingrowth and individual species regeneration niches

```{r nvTrend710BAlightTemp, echo=FALSE, fig.cap="Mean ingrowth values across the 200 samples per site, for the 200 sites against gradients of (a) total basal area; (b) climatic water balance; (c) seasonal degree sum. The values were split into ten bins; the red lines represent a GAM model showing the trend in the observed data. The same graphs with y-axis scaled by model as availalbe in Figure S", fig.show = 'hold', out.width="49%",fig.ncol = 2, fig.subcap=c('', '', ''), fig.show='hold', fig.align='left'}
     knitr::include_graphics(c(here::here("figures", "envTrend7Totba.png"),
                              here::here("figures", "envTrend7WB.png"),
                              here::here("figures", "envTrend7DDS.png")))
```
